---
title: "Using tenzing as a package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{local_use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

If you do not want to use the tenzing app, you can use the package from R to achieve the same outputs.

# Setup

First, you have to install the package.

```{r, eval = FALSE}
# install.packages("devtools")
 devtools::install_github("marton-balazs-kovacs/tenzing")
```

Second you have to load the package.

```{r, message = FALSE, warning = FALSE}
library(tenzing)
```

# Create your infosheet

The infosheet template is built in the package, as well as uploaded to the [net](https://docs.google.com/spreadsheets/d/1Gl0cwqN_nTsdFH9yhSvi9NypBfDCEhViGq4A3MnBrG8/edit?usp=sharing). If you choose to fill out the template with your credit information locally, you can write the infosheet template as an xlsx file to your working directory from the package with the following code:

```{r, eval = FALSE}
# install.packages("writexl")
writexl::write_xlsx(infosheet_template, "my_infosheet.xlsx")
```

To get more information on the infosheet tempalte use the `?tenzing::infosheet_template` command.

_Note: This produces the same result as downloading the infosheet template from the link provided before._

# Load your infosheet

You can load the infosheet into R with the `tenzing::read_infosheet` function. This function accepts files with csv, tsv, xlsx extensions, and the share URL of the Google spreadhseet.
As an example we will use the built in infosheet template.

```{r}
file_path <- system.file("extdata", "infosheet_template_example.csv", package = "tenzing", mustWork = TRUE)
my_infosheet <- read_infosheet(infosheet_path = file_path)
```

If the read infosheet still contains empty rows you can clean it with the `clean_infosheet` function.

```{r}
my_infosheet <- clean_infosheet(my_infosheet)
```

After loading your infosheet you have to check its validity.

```{r}
check_result <- validate_infosheet(infosheet = my_infosheet)
```

The validation function checks your infosheet by 10 statements. You can learn more about these checks by running the following code `?tenzing::infosheet_template`.

The check can have three results: _success_, _warning_ or _error_. You can investigate what was the output of each of the 10 checks.

```{r}
purrr::map(check_result, "type")
```

If any of them is an _error_ than the output that you will generate in the next step will be corroborated. Therefore, you have to read the messages that are accompanied with the erroneous checks and update your infosheet accordingly.

```{r}
purrr::map(check_result, "message")
```

# Generate output

If your infosheet is validated you can move on to output generation. There are four different types of outputs that you can create with the tenzing app.

For the human readable report and the contributors' affiliation page the output text will be rmarkdown formatted by default. However, by setting the `text_format` argument to `"html"` the output can be HTML formatted as well, or by setting the argument to `"raw"`, the output string will not be formatted at all.

## create a human readable report of the contributions according to the CRediT taxonomy

For this section it is possible to use initials by setting the `initials` argument `TRUE`. Also, if the `order_by` argument is set to `contributor` the function will list the contributions after the name of each researcher, instead of listing the appropriate names after each CRediT role.

```{r}
print_credit_roles(infosheet = my_infosheet, initials = TRUE, order_by = "contributor")
```

## Create the contributors' affiliation page

```{r}
print_title_page(infosheet = my_infosheet)
```

## Create a JATS formatted XML document containing the contributors information

```{r}
print_xml(infosheet = my_infosheet)
```

## Create a YAML document containing the contributors informtation

This output can be incorporated into manuscript created with the `papaja` package.

```{r}
print_yaml(infosheet = my_infosheet)
```

## Create funding acknowledgements section

For this section it is possible to use initials by setting the `initials` argument `TRUE`.

```{r}
print_funding(infosheet = my_infosheet, initials = TRUE)
```

