---
title: "Using tenzing as a package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{local_use}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

If you do not want to use the __tenzing__ app, you can use the package from R to achieve the same outputs.

# Setup

First, you have to install the package.

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("marton-balazs-kovacs/tenzing")
```

Second you have to load the package.

```{r, message = FALSE, warning = FALSE}
library(tenzing)
```

# Create your contributors table

The contributors table template is built in the package, as well as uploaded to the [net](https://docs.google.com/spreadsheets/d/1ay8pS-ftvfzWTrKCZr6Fa0cTLg3n8KxAOOleZmuE7Hs/edit?usp=sharing). If you choose to fill out the template with your CRediT information locally, you can write your contributors table as an xlsx file to your working directory from the package with the following code:

```{r, eval = FALSE}
# install.packages("writexl")
writexl::write_xlsx(contributors_table_template, "my_contributors_table.xlsx")
```

To get more information on the contributors table template use the `?tenzing::contributors_table_template` command.

_Note: This produces the same result as downloading the contributors table template from the link provided before._

_Note: The contributors table template was changed since the first release, as some of the CRediT roles were not named properly in the template columns._

# Load your contributors table

You can load the contributors table into R with the `tenzing::read_contributors_table` function. This function accepts files with csv, tsv, xlsx extensions, and the share URL of the Google spreadhseet.
As an example we will use the built in contributors table template.

```{r}
file_path <- system.file("extdata", "contributors_table_example.csv", package = "tenzing", mustWork = TRUE)
my_contributors_table <- read_contributors_table(contributors_table_path = file_path)
```

If the read contributors table still contains empty rows you can clean it with the `clean_contributors_table` function.

```{r}
my_contributors_table <- clean_contributors_table(my_contributors_table)
```

# Validate your contributors table

Before generating outputs, you must check whether your `contributors_table` is well-formatted. In __tenzing__, validation is handled using two R6 classes:

* `ColumnValidator`: Ensures required columns exist.
* `Validator`: Runs logical and consistency checks on the data.

Both work based on __YAML configuration files__, allowing for flexible, customizable validation pipelines. For a detailed explanation of validation rules and configurations, please refer to the `Validation Vignette`.

__tenzing__ supports multiple output types, each requiring a properly formatted `contributors_table.` You can use the `ValidateOutput` R6 class to check your table __against a specific configuration__ for each output.

For example, if you want to generate a __title page__ for your manuscript, you should first validate your table to ensure it meets the required structure.

## 1. Load the Title Page Validation Configuration

```{r}
config_path <- system.file("config/title_validation.yaml", package = "tenzing")
```

## 2. Initialize a Validation Instance

```{r}
validate_output_instance <- ValidateOutput$new(config_path = config_path)
```

## 3. Run the Validation Checks

```{r}
validate_results <- validate_output_instance$run_validations(contributors_table = my_contributors_table)
```

## 4. Review Validation Results

Each check returns one of three statuses:

* __success__ – The check passed.
* __warning__ – The table has issues that may affect output quality.
* __error__ – The table is incorrectly formatted and must be fixed before generating output. 

You can check the status of each validation:

```{r}
purrr::map(validate_results, "type")
```

And review the detailed messages for failed checks:

```{r}
purrr::map(validate_results, "message")
```

# Generate output

If your contributors table is validated you can move on to output generation. There are six different types of outputs that you can create with the __tenzing__ app.

For the human readable report and the contributors' affiliation page the output text will be rmarkdown formatted by default. However, by setting the `text_format` argument to `"html"` the output can be HTML formatted as well, or by setting the argument to `"raw"`, the output string will not be formatted at all.

## create a human readable report of the contributions according to the CRediT taxonomy

For this section it is possible to use initials by setting the `initials` argument `TRUE`. Also, if the `order_by` argument is set to `contributor` the function will list the contributions after the name of each researcher, instead of listing the appropriate names after each CRediT role.

```{r}
print_credit_roles(contributors_table = my_contributors_table, initials = TRUE, order_by = "contributor")
```

## Create the contributors' affiliation page

```{r}
print_title_page(contributors_table = my_contributors_table)
```

## Create a JATS formatted XML document containing the contributors information

```{r}
print_xml(contributors_table = my_contributors_table)
```

## Create a YAML document containing the contributors informtation

This output can be incorporated into manuscript created with the `papaja` package.

```{r}
print_yaml(contributors_table = my_contributors_table)
```

## Create funding acknowledgements section

For this section it is possible to use initials by setting the `initials` argument `TRUE`.

```{r}
print_funding(contributors_table = my_contributors_table, initials = TRUE)
```

## Create a conflict of interest statement

For this section it is possible to use the initials by setting the `initials` argument `TRUE`.

```{r}
print_conflict_statement(contributors_table = my_contributors_table, initials = FALSE)
```

