---
title: "Custom Validation in tenzing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{validation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE, message = FALSE, warning = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tenzing)

file_path <- system.file("extdata", "contributors_table_example.csv", package = "tenzing", mustWork = TRUE)
my_contributors_table <- read_contributors_table(contributors_table_path = file_path)
```

This vignette provides a step-by-step guide on how to create and apply
custom validation rules in **tenzing**. The validation framework in
**tenzing** is based on R6 classes that allow for flexible and
configurable validation of contributor tables.

# Overview

The validation system in **tenzing** consists of three main components:

-   `ColumnValidator` -- Ensures that required columns exist in the
    contributors table.
-   `Validator` -- Runs logical checks on the contents of the table
    (e.g., missing values, duplicate names).
-   `ValidateOutput` -- Combines column and data validation, allowing
    for customized validation pipelines using `YAML` configuration
    files.

By leveraging these components, you can create your own validation rules
to check for specific issues in your data.

# 1. Defining Custom Validation Rules

Validation rules are written as functions in **tenzing**. These
functions should take the `contributors_table` as input and return a
list with two elements:

-   `type`: Can be "success", "warning", or "error", indicating the
    result of the check.
-   `message`: A user-friendly explanation of the check result.

# Example: Custom Validation Function

Let's say you want to create a check that ensures every contributor has
a valid ORCID ID.

```{r}
#' Check for valid ORCID IDs
#'
#' This function checks if the ORCID IDs in the `contributors_table` are formatted correctly.
#'
#' @param contributors_table A dataframe containing the contributors' information.
#'
#' @return A list containing:
#' \item{type}{Type of validation result: "success", "warning", or "error".}
#' \item{message}{An informative message indicating if any ORCID IDs are invalid.}
check_orcid <- function(contributors_table) {
  if (!"ORCID" %in% colnames(contributors_table)) {
    return(list(
      type = "warning",
      message = "No ORCID column found. ORCID validation skipped."
    ))
  }
  
  invalid_orcids <- contributors_table %>%
    dplyr::filter(!grepl("^\\d{4}-\\d{4}-\\d{4}-\\d{4}$", .data$ORCID) & !is.na(.data$ORCID))
  
  if (nrow(invalid_orcids) > 0) {
    return(list(
      type = "warning",
      message = glue::glue("Invalid ORCID format for the following rows: {paste(invalid_orcids$rowname, collapse = ', ')}")
    ))
  }
  
  return(list(type = "success", message = "All ORCID IDs are correctly formatted."))
}

```

# 2. Configuring Validation with YAML

Once you define your validation function, you need to tell **tenzing**
to use it. This is done by specifying the validation in a `YAML`
configuration file.

**tenzing** includes a set of predefined validation helpers in `R/validation_helpers.R`. These include:
- `check_missing_order` – Ensures that all contributors have an order in the publication.
- `check_duplicate_order` – Ensures no duplicate order numbers unless multiple first authors exist.
- `check_missing_surname` – Ensures all contributors have a surname.
- `check_duplicate_names` – Ensures no duplicate contributor names.
- `check_affiliation_consistency` – Ensures only one affiliation format is used.
- Many more...

These functions are automatically available when setting up validation in **tenzing**.

## Writing Custom Dependencies

Some validations should only run if other conditions are met. You can define dependencies in the `YAML` configuration file.

### Example: Adding ORCID Validation in `config/validator_vignette_example.yaml`

```         
validation_config:
  validations:
    - name: check_missing_order
    - name: check_duplicate_order
    - name: check_missing_surname
    - name: check_missing_firstname
    - name: check_duplicate_initials
    - name: check_missing_corresponding
      dependencies:
        - '"Corresponding author?" %in% colnames(contributors_table)'
    - name: check_missing_email
      dependencies:
        - '"Corresponding author?" %in% colnames(contributors_table)'
        - 'self$results[["check_missing_corresponding"]]$type == "success"'
        - '"Email address" %in% colnames(contributors_table)'
    - name: check_orcid
      dependencies:
        - '"ORCID" %in% colnames(contributors_table)'
```

In this configuration:

-   The `check_orcid` validation will only run if the column `"ORCID"` exists.
-   The `check_missing_email` validation will only run if a corresponding author is specified and their email column exists.

# 3. Running Custom Validations

Once you have added the validation rule, you can run it using **tenzing's** validation pipeline.

## Step 1: Load the Configuration

```{r}
config_path <- system.file("config/validator_vignette_example.yaml", package = "tenzing")

config_file <- yaml::read_yaml(config_path)
```

## Step 2: Initialize the `Validator` class

```{r}
validator <- Validator$new()

validator$setup_validator(config_file$validation_config)
```

## Step 3: Run Validations on Your Data

```{r}
validate_results <- validator$run_validations(contributors_table = my_contributors_table)
```

## Step 4: Inspect the Validation Results

```{r}
purrr::map(validate_results, "type")
```

```{r}
purrr::map(validate_results, "message")
```

# 4. Understanding the `Validator` Class

The `Validator` class in **tenzing** is responsible for running all
validation checks.

Key Features of `Validator`

-   It dynamically loads validation functions from
    `validation_helpers.R`.
-   It allows adding dependencies between validation rules.
-   It executes only the specified validations from the configuration.

# 5. Ensuring Required Columns Exist with ColumnValidator

The `ColumnValidator` class ensures that all necessary columns are present before running validations.

## Example: Configuring Required Columns

In `config/columnvalidator_example.yaml`:

```         
column_config:
  rules:
    minimal:
      operator: "AND"
      columns:
        - Firstname
        - Middle name
        - Surname
        - Order in publication
      severity: "error"
    affiliation:
      operator: "OR"  # Either legacy OR regex-based affiliation columns must be present
      columns:
        - Primary affiliation
        - Secondary affiliation  # Legacy columns
      regex: "^Affiliation [0-9]+$"  # Regex-based columns
      severity: "error"  # Make sure it's required for validation to pass
    title:
      operator: "AND"
      columns:
        - Corresponding author?
        - Email address
      severity: "warning"
```

## Using Regex to Validate Column Names

You can also define **regex-based column validation** to match dynamically named columns.

### Example: Using Regex for Affiliation Columns

```
column_config:
  rules:
    affiliation:
      operator: "OR"
      columns:
        - Primary affiliation
        - Secondary affiliation
      regex: "^Affiliation [0-9]+$"
      severity: "error"
```

This configuration:
-   Requires at least one affiliation column.
-   Allows dynamic affiliation columns (e.g., `Affiliation 1`, `Affiliation 2`, etc.).
-   Fails validation if no affiliation column exists.


## Running the ColumnValidator

```{r}
config_path <- system.file("config/columnvalidator_example.yaml", package = "tenzing")

config_file <- yaml::read_yaml(config_path)

column_validator <- ColumnValidator$new(config_input = config_file$column_config)

column_results <- column_validator$validate_columns(my_contributors_table)

column_results
```

# 6. Bringing It All Together with ValidateOutput

The `ValidateOutput` class integrates both `Validator` and
`ColumnValidator`.

## How It Works

1.  Reads the configuration file.

2.  Runs ColumnValidator to check required columns.

3.  Runs Validator to check data integrity.

4.  Returns the combined results.

## Example Usage

```{r}
validate_output_instance <- ValidateOutput$new(config_path = config_path)

validate_results <- validate_output_instance$run_validations(my_contributors_table)

validate_results
```

# 7. Summary

-   Define custom validation functions in `validation_helpers.R`.
-   Add custom validations to `YAML` configuration files.
-   Use `Validator` to run content-based checks.
-   Use `ColumnValidator` to enforce required columns.
-   Use `ValidateOutput` to integrate everything.
