---
title: "Custom Validation in tenzing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{validation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include = FALSE, message = FALSE, warning = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tenzing)

file_path <- system.file("extdata", "contributors_table_example.csv", package = "tenzing", mustWork = TRUE)
my_contributors_table <- read_contributors_table(contributors_table_path = file_path)
```

This vignette provides a step-by-step guide on how to create and apply
custom validation rules in **tenzing**. The validation framework in
**tenzing** is based on R6 classes that allow for flexible and
configurable validation of contributor tables.

# Overview

The validation system in **tenzing** consists of three main components:

-   `ColumnValidator` -- Ensures that required columns exist in the
    contributors table.
-   `Validator` -- Runs logical checks on the contents of the table
    (e.g., missing values, duplicate names).
-   `ValidateOutput` -- Combines column and data validation, allowing
    for customized validation pipelines using `YAML` configuration
    files.

By leveraging these components, you can create your own validation rules
to check for specific issues in your data.

# 1. Defining Custom Validation Rules

Validation rules are written as functions in **tenzing**. These
functions should take the `contributors_table` as input and return a
list with two elements:

-   `type`: Can be "success", "warning", or "error", indicating the
    result of the check.
-   `message`: A user-friendly explanation of the check result.

# Example: Custom Validation Function

Let's say you want to create a check that ensures every contributor has
a valid ORCID ID.

```{r}
#' Check for valid ORCID IDs
#'
#' This function checks if the ORCID IDs in the `contributors_table` are formatted correctly.
#'
#' @param contributors_table A dataframe containing the contributors' information.
#'
#' @return A list containing:
#' \item{type}{Type of validation result: "success", "warning", or "error".}
#' \item{message}{An informative message indicating if any ORCID IDs are invalid.}
check_orcid <- function(contributors_table) {
  if (!"ORCID" %in% colnames(contributors_table)) {
    return(list(
      type = "warning",
      message = "No ORCID column found. ORCID validation skipped."
    ))
  }
  
  invalid_orcids <- contributors_table %>%
    dplyr::filter(!grepl("^\\d{4}-\\d{4}-\\d{4}-\\d{4}$", .data$ORCID) & !is.na(.data$ORCID))
  
  if (nrow(invalid_orcids) > 0) {
    return(list(
      type = "warning",
      message = glue::glue("Invalid ORCID format for the following rows: {paste(invalid_orcids$rowname, collapse = ', ')}")
    ))
  }
  
  return(list(type = "success", message = "All ORCID IDs are correctly formatted."))
}

```

# 2. Understanding the Configuration System

**tenzing** uses a clean, consistent configuration system where all validation files follow the same structure:

-   **General column validation**: `column_validation.yaml` for comprehensive column checking
-   **Output-specific validation**: Files like `title_validation.yaml` for specific output types  
-   **Consistent structure**: All configs use `column_config` and `validation_config` sections
-   **Easy customization**: You can create custom configurations or extend existing ones

**tenzing** includes a set of predefined validation helpers in `R/validate_helpers.R`. These include:
- `check_missing_order` – Ensures that all contributors have an order in the publication.
- `check_duplicate_order` – Ensures no duplicate order numbers unless multiple first authors exist.
- `check_missing_surname` – Ensures all contributors have a surname.
- `check_duplicate_names` – Ensures no duplicate contributor names.
- `check_affiliation_consistency` – Ensures only one affiliation format is used.
- Many more...

These functions are automatically available when setting up validation in **tenzing**.

## Understanding General vs Output-Specific Validation

**tenzing** uses two types of validation:

1. **General Column Validation** (`column_validation.yaml`):
   - Checks all possible columns that might be needed across different outputs
   - Runs before any output-specific validation

2. **Output-Specific Validation** (e.g., `title_validation.yaml`):
   - Checks columns and data specific to a particular output type
   - Used by `ValidateOutput` class for detailed validation
   - Runs after general column validation passes

## Using Predefined Configuration Files

**tenzing** provides ready-to-use configuration files for common output types:

```{r}
# List available configuration files
config_dir <- system.file("config", package = "tenzing")
list.files(config_dir, pattern = "*_validation\\.yaml$")
```

Each configuration file specifies both:
1. **Column validation rules**: Which columns are required
2. **Data validation functions**: Which checks to perform

### Example: Using the Title Page Configuration

The `title_validation.yaml` file includes validation for title page outputs:

```{r, eval = FALSE}
# Load the title validation configuration
title_config_path <- system.file("config/title_validation.yaml", package = "tenzing")
title_config <- yaml::read_yaml(title_config_path)

# View the column rules
title_config$column_config$rules

# View the validation functions
purrr::map_chr(title_config$validation_config$validations, "name")
```

## Creating Custom Configurations

You can create your own validation configuration file with custom rules and validations.

### Example: Creating a Custom Configuration with ORCID Validation

First, create a YAML configuration file (e.g., `my_custom_validation.yaml`):

```         
column_config:
  rules:
    minimal_rule:
      operator: "AND"
      columns:
        - Firstname
        - Middle name
        - Surname
        - Order in publication
      severity: "error"
    orcid_rule:
      operator: "AND"
      columns:
        - ORCID
      severity: "warning"

validation_config:
  validations:
    - name: check_missing_order
    - name: check_duplicate_order
    - name: check_missing_surname
    - name: check_missing_firstname
    - name: check_duplicate_initials
    - name: check_duplicate_names
    - name: check_orcid
      dependencies:
        - '"ORCID" %in% colnames(contributors_table)'
```

### Writing Custom Dependencies

Some validations should only run if other conditions are met. You can define dependencies in the `YAML` configuration file.

In the example above:
-   The `check_orcid` validation will only run if the column `"ORCID"` exists.
-   Dependencies can check for column existence, previous validation results, or context variables.

# 3. Running Validations

You can run validations using **tenzing's** validation pipeline in two ways:

## Method 1: Using ValidateOutput (Recommended)

The `ValidateOutput` class is the easiest way to run validations, as it handles both column and data validation automatically.

### Step 1: Initialize ValidateOutput with a Configuration

```{r}
# Use a predefined configuration
config_path <- system.file("config/title_validation.yaml", package = "tenzing")

validate_output <- ValidateOutput$new(config_path = config_path)
```

### Step 2: Run Validations

```{r}
validate_results <- validate_output$run_validations(my_contributors_table)
```

### Step 3: Inspect the Validation Results

```{r}
# View validation types
purrr::map_chr(validate_results, "type")
```

```{r}
# View validation messages
purrr::map_chr(validate_results, "message")
```

## Method 2: Using Validator Directly

For more control, you can use the `Validator` class directly.

### Step 1: Load the Configuration

```{r}
config_path <- system.file("config/title_validation.yaml", package = "tenzing")
config_file <- yaml::read_yaml(config_path)
```

### Step 2: Initialize the `Validator` class

```{r}
validator <- Validator$new()

validator$setup_validator(config_file$validation_config)
```

### Step 3: Run Validations on Your Data

```{r}
validate_results <- validator$run_validations(contributors_table = my_contributors_table)
```

### Step 4: Inspect the Validation Results

```{r}
purrr::map_chr(validate_results, "type")
```

```{r}
purrr::map_chr(validate_results, "message")
```

# 4. Understanding the Validation Classes

## The `Validator` Class

The `Validator` class in **tenzing** is responsible for running all
data validation checks.

Key Features of `Validator`:

-   It loads validation functions explicitly from
    `validate_helpers.R`, ensuring they work in all environments.
-   It allows adding dependencies between validation rules.
-   It executes only the specified validations from the configuration.
-   It supports context-aware validation for dynamic UI states.

## The `ColumnValidator` Class

The `ColumnValidator` class ensures that all necessary columns are present before running validations.

Key Features of `ColumnValidator`:

-   Supports logical operators: `AND`, `OR`, `NOT` for column requirements.
-   Supports regex patterns for dynamically named columns.
-   Supports severity levels: `error` or `warning`.

## The `ValidateOutput` Class

The `ValidateOutput` class integrates both `ColumnValidator` and `Validator` into a single, easy-to-use interface.

Key Features of `ValidateOutput`:

-   Automatically runs column validation before data validation.
-   Returns only column validation errors if critical columns are missing.
-   Supports optional context parameters for dynamic validation.

# 5. Column Validation Examples

The `ColumnValidator` class ensures that all necessary columns are present before running validations.

## Example: Understanding Column Validation Rules

Let's examine the column validation rules in the title configuration:

```{r}
config_path <- system.file("config/title_validation.yaml", package = "tenzing")
config <- yaml::read_yaml(config_path)

# View the column rules
config$column_config$rules
```

The configuration defines several rules:

1. **`minimal_rule`**: Requires basic contributor information (Firstname, Middle name, Surname, Order in publication) with `AND` operator (all must be present).

2. **`affiliation_rule`**: Requires at least one affiliation column (using `OR` operator) and supports both legacy columns (`Primary affiliation`, `Secondary affiliation`) and regex-based columns matching `^Affiliation [0-9]+$`.

3. **`title_rule`**: Requires corresponding author information (Both `Corresponding author?` and `Email address` must be present).

## Using Regex to Validate Column Names

You can define **regex-based column validation** to match dynamically named columns, which is useful for scenarios like multiple affiliation columns.

### Example: Regex for Affiliation Columns

The affiliation rule uses both legacy columns and regex:

```yaml
affiliation_rule:
  operator: "OR"
  columns:
    - Primary affiliation
    - Secondary affiliation
  regex: "^Affiliation [0-9]+$"
  severity: "error"
```

This configuration:
-   Requires at least one affiliation column.
-   Matches legacy columns (`Primary affiliation`, `Secondary affiliation`).
-   Matches dynamic columns (e.g., `Affiliation 1`, `Affiliation 2`, etc.) using regex.
-   Fails validation if no affiliation column exists.

## Running ColumnValidator Independently

You can run column validation independently to check column requirements:

```{r}
config_path <- system.file("config/title_validation.yaml", package = "tenzing")
config <- yaml::read_yaml(config_path)

column_validator <- ColumnValidator$new(config_input = config$column_config)

column_results <- column_validator$validate_columns(my_contributors_table)

# View results
purrr::map_chr(column_results, "type")
```

```{r}
purrr::map_chr(column_results, "message")
```

# 6. Using ValidateOutput for Complete Validation

The `ValidateOutput` class integrates both `ColumnValidator` and
`Validator` into a unified validation pipeline.

## How ValidateOutput Works

1.  Reads the configuration file (either from a path or with base config merging).

2.  Runs `ColumnValidator` to check required columns (stops early if critical columns are missing).

3.  Runs `Validator` to check data integrity (only if column validation passes).

4.  Returns the combined results from both validation stages.

## Example: Using Predefined Configurations

```{r}
# Title page validation
title_config <- system.file("config/title_validation.yaml", package = "tenzing")
validate_output <- ValidateOutput$new(config_path = title_config)

validate_results <- validate_output$run_validations(my_contributors_table)

# Check for any errors
has_errors <- any(purrr::map_chr(validate_results, "type") == "error")
has_errors
```

## Example: Using Context-Aware Validation

Some validations can use context information (e.g., user selections in a Shiny app):

```{r, eval = FALSE}
# Create context for filtering
context <- list(include = "author", order_by = "contributor", pub_order = "asc")

# Run validation with context
validate_results <- validate_output$run_validations(
  my_contributors_table, 
  context = context
)
```

The context allows validations to react to dynamic UI states, such as:
-   Filtering by author vs. acknowledgee
-   Different ordering preferences
-   User-selected output options

# 7. Configuration Management

## Configuration Utilities

**tenzing** provides utility functions for configuration management:

```{r, eval = FALSE}
# Clear configuration cache
clear_config_cache()

# Get cache statistics
get_cache_stats()

# Validate configuration schema
validate_config_schema(config)
```

## Best Practices

1. **Start with predefined configurations**: Use `title_validation.yaml`, `credit_validation.yaml`, etc., as starting points.

2. **Use the two-tier approach**: Run general column validation first, then output-specific validation.

3. **Use dependencies wisely**: Only add dependencies when validations truly depend on each other or column presence.

4. **Test your configurations**: Ensure your custom validations work correctly before using them in production.

# 8. Summary

-   **Define custom validation functions** that return lists with `type` and `message`.
-   **Use the two-tier validation approach**: General column validation first, then output-specific validation.
-   **Use predefined configuration files** for common output types (title, credit, yaml, etc.).
-   **Create custom YAML configurations** following the consistent `column_config` and `validation_config` structure.
-   **Use `ColumnValidator`** to enforce required columns before data validation.
-   **Use `Validator`** to run content-based validation checks.
-   **Use `ValidateOutput`** for the easiest integration of column and data validation.
-   **Leverage the clean configuration system** for consistency and maintainability.
